Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-03-13 r-journal/manuscript.tex
update paper

% !TeX root = RJwrapper.tex
\title{Partitioned Local Depth (PaLD) Clustering Analyses in R}
\author{by Lucy D'Agostino McGowan, Katherine Moore, Kenneth Berenhaut}

@@

% Any extra LaTeX you need in the preamble

@@

\begin{itemize}
\tightlist
\item
  Describe PaLD (Ken \& Kate?)
\item
  Cite PaLD
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Helper functions to organize data into the correct format, into
  distance matrices and then contribution matrices
\item
  Functions that convert a contribution matrix into a variety of useful
  formats, including partitioned local depths, clusters, and graphs
\item
  Plotting functions
\end{enumerate}

@@

While it is not a necessity, the \CRANpkg{pald} package is designed to
function well with the pipe operator, \texttt{\%\textgreater{}\%}, from
the \citet{magrittr} \CRANpkg{magrittr} package in that the first
argument of each function is the data. This functionality will be
demonstrated below.

@@

For demonstration purposes, below is a sample data frame with two
variables, \texttt{x1} and \texttt{x2}. The methods put forth here work
on data frames with higher dimensions, as described in the
\textbf{Examples} section, we are simply choosing a small data frame
here for demonstration purposes.

\begin{Schunk}
\begin{Sinput}
library(pald)
d <- data.frame(
  x1 = c(6, 8, 11, 16, 4),
  x2 = c(5, 4, 13, 7, 18)
)
rownames(d) <- c("A", "B", "C", "D", "E")
\end{Sinput}
\end{Schunk}

The first step needed to calculate the partitioned local depths is to
construct a \emph{distance matrix}. If the data are already in this
form, the user can skip to the next step. The
\texttt{get\_distance\_matrix()} function converts an input data frame
into a distance matrix, as demonstrated below.

\begin{Schunk}
\begin{Sinput}
get_distance_matrix(d)
\end{Sinput}
\begin{Soutput}
#>           A         B        C        D        E
#> A 0.0000000 0.4584177 1.717380 2.158418 2.229205
#> B 0.4584177 0.0000000 1.644296 1.778777 2.505929
#> C 1.7173802 1.6442957 0.000000 1.468398 1.713327
#> D 2.1584182 1.7787772 1.468398 0.000000 3.158040
#> E 2.2292047 2.5059291 1.713327 3.158040 0.000000
\end{Soutput}
\end{Schunk}

This will create an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 5\). By default, the distance matrix is
\emph{scaled}, is possible to create a distance matrix that is not
scaled by changing the \texttt{scale} argument to \texttt{FALSE}.

\begin{Schunk}
\begin{Sinput}
get_distance_matrix(d, scale = FALSE)
\end{Sinput}
\begin{Soutput}
#>           A         B        C         D         E
#> A  0.000000  2.236068 9.433981 10.198039 13.152946
#> B  2.236068  0.000000 9.486833  8.544004 14.560220
#> C  9.433981  9.486833 0.000000  7.810250  8.602325
#> D 10.198039  8.544004 7.810250  0.000000 16.278821
#> E 13.152946 14.560220 8.602325 16.278821  0.000000
\end{Soutput}
\end{Schunk}

As mentioned previously, the functions in this package are designed to
work with the \CRANpkg{magrittr} pipe operator,
\texttt{\%\textgreater{}\%}, so the same code as above could be written
utilizing this format.

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix()
\end{Sinput}
\begin{Soutput}
#>           A         B        C        D        E
#> A 0.0000000 0.4584177 1.717380 2.158418 2.229205
#> B 0.4584177 0.0000000 1.644296 1.778777 2.505929
#> C 1.7173802 1.6442957 0.000000 1.468398 1.713327
#> D 2.1584182 1.7787772 1.468398 0.000000 3.158040
#> E 2.2292047 2.5059291 1.713327 3.158040 0.000000
\end{Soutput}
\end{Schunk}

This distance matrix can then be passed to the
\texttt{get\_contribution\_matrix()} function in order to calculate a
matrix of contributions. Again, if the user begins with a distance
matrix, they can skip the first step and simple input the contribution
matrix into this function.

\begin{Schunk}
\begin{Sinput}
distance_matrix <- get_distance_matrix(d)
get_contribution_matrix(distance_matrix)
\end{Sinput}
\begin{Soutput}
#>        1      2     3     4      5
#> A 0.2875 0.1625 0.000 0.050 0.0000
#> B 0.1750 0.3000 0.050 0.050 0.0000
#> C 0.0000 0.0625 0.300 0.175 0.0500
#> D 0.0500 0.0500 0.175 0.300 0.0000
#> E 0.0000 0.0000 0.050 0.000 0.2125
\end{Soutput}
\end{Schunk}

Again, the \CRANpkg{magrittr} pipe can be used as follows.

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
\end{Sinput}
\begin{Soutput}
#>        1      2     3     4      5
#> A 0.2875 0.1625 0.000 0.050 0.0000
#> B 0.1750 0.3000 0.050 0.050 0.0000
#> C 0.0000 0.0625 0.300 0.175 0.0500
#> D 0.0500 0.0500 0.175 0.300 0.0000
#> E 0.0000 0.0000 0.050 0.000 0.2125
\end{Soutput}
\end{Schunk}

The \emph{contribution matrix} output by the
\texttt{get\_contribution\_matrix()} is the main input for the majority
of the remaining functions.

\hypertarget{functions-that-convert-a-contribution-matrix-into-useful-formats}{%
\subsubsection{Functions that convert a contribution matrix into useful
formats}\label{functions-that-convert-a-contribution-matrix-into-useful-formats}}

From the \emph{contribution matrix}, a variety of useful quantities can
be calculated. Below, we create a contribution matrix using the
functions described in the previous section.

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() -> contribution_matrix
\end{Sinput}
\end{Schunk}

To calculate the \emph{clusters} that each point will fall into, we can
use the \texttt{get\_clusters()} function. This will output a data frame
with two columns, the first will correspond to the \texttt{point}, as
identified by the row name of the original input data frame, \texttt{d},
the second will identify the \texttt{cluster} that each point belongs
to.

\begin{Schunk}
\begin{Sinput}
get_clusters(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#> # A tibble: 5 x 2
#>   point cluster
#>   <chr>   <dbl>
#> 1 A           1
#> 2 B           1
#> 3 C           2
#> 4 D           2
#> 5 E           3
\end{Soutput}
\end{Schunk}

@@

The \texttt{get\_depths()} function calculates the \emph{depths} of each
point, outputting a data frame with two columns, \texttt{point}
indicating the point, as identified by the row name of the original data
frame, \texttt{d}, and \texttt{depth} indicating the depth of the point.
The data frame is arranged by depth, with the deepest point listed
first.

\begin{Schunk}
\begin{Sinput}
get_depths(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#> # A tibble: 5 x 2
#>   point depth
#>   <chr> <dbl>
#> 1 C     0.588
#> 2 B     0.575
#> 3 D     0.575
#> 4 A     0.5  
#> 5 E     0.262
\end{Soutput}
\end{Schunk}

@@

The \texttt{get\_bound()} function will calculate the \emph{bound} of
the contribution matrix.

\begin{Schunk}
\begin{Sinput}
get_bound(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#> [1] 0.14
\end{Soutput}
\end{Schunk}

In this case, the bound is 0.14.

@@

\begin{Schunk}
\begin{Sinput}
any_isolated(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#> [1] FALSE
\end{Soutput}
\end{Schunk}

In this case, there are no isolated points.

The \texttt{get\_pald\_cluster\_matrix()} will calculate a matrix of
partitioned local depth clusters. This function contains an argument
\texttt{keep\_all\_edges} that indicates whether all edges should be
kept. The default value is \texttt{FALSE}, indicating that edges that
are less than the expectation will be dropped, allowing a clear
clustering of points. Changing this argument to \texttt{TRUE} results in
the pairwise minimum of the contribution matrix and the transpose of the
contribution matrix.

\begin{Schunk}
\begin{Sinput}
get_pald_cluster_matrix(contribution_matrix, keep_all_edges = TRUE)
\end{Sinput}
\begin{Soutput}
#>        1      2     3     4      5
#> A 0.2875 0.1625 0.000 0.050 0.0000
#> B 0.1625 0.3000 0.050 0.050 0.0000
#> C 0.0000 0.0500 0.300 0.175 0.0500
#> D 0.0500 0.0500 0.175 0.300 0.0000
#> E 0.0000 0.0000 0.050 0.000 0.2125
\end{Soutput}
\end{Schunk}

Leaving the \texttt{keep\_all\_edges} as the default, \texttt{FALSE}
will take this pairwise minimum seen above and drop the edges that are
less than the expectation.

\begin{Schunk}
\begin{Sinput}
get_pald_cluster_matrix(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#>        1      2     3     4 5
#> A 0.0000 0.1625 0.000 0.000 0
#> B 0.1625 0.0000 0.000 0.000 0
#> C 0.0000 0.0000 0.000 0.175 0
#> D 0.0000 0.0000 0.175 0.000 0
#> E 0.0000 0.0000 0.000 0.000 0
\end{Soutput}
\end{Schunk}

Finally, the \texttt{get\_pald\_graph()} function takes the contribution
matrix and creates an \CRANpkg{igraph} object, a graph that describes
the relationship between the points.

\begin{Schunk}
\begin{Sinput}
get_pald_graph(contribution_matrix)
\end{Sinput}
\begin{Soutput}
#> IGRAPH 2c0b54a UNW- 5 2 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 2c0b54a (vertex names):
#> [1] A--B D--C
\end{Soutput}
\end{Schunk}

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{d} (Figure
\ref{fig:fig1}).

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(d, aes(x1, x2)) +
  geom_text(label = rownames(d))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `d`]{Visualize the points from data frame `d`}\label{fig:fig1}
\end{figure}
\end{Schunk}

We can then pass the contribution matrix to the
\texttt{plot\_pald\_graph()} function to view the relationship between
points (Figure \ref{fig:fig2}). By default, this will invoke the
\texttt{layout\_nicely()} function from \CRANpkg{igraph} to determine
the layout of the graph.

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph()
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`]{PaLD graph displaying the relationship between the points in data frame `d`}\label{fig:fig2}
\end{figure}
\end{Schunk}

The \texttt{layout} argument allows the user to pass a matrix to dictate
the layout of the graph. For example, if we wanted the graph to match
the visualization displayed in Figure \ref{fig:fig1}, we can pass
\texttt{as.matrix(d)}, or a matrix of the data frame \texttt{d} to the
\texttt{layout} argument (Figure \ref{fig:fig3}.

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph(layout = as.matrix(d))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

This \texttt{plot\_pald\_graph()} function will also permit parameters
that can be passed to the \texttt{plot.igraph()} function. For example,
to add axes to the graph, the user can pass the \texttt{axes\ =\ TRUE}
argument to the \texttt{...} in the \texttt{plot\_pald\_graph()}
function (Figure \ref{fig:fig4}).

\begin{Schunk}
\begin{Sinput}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph(layout = as.matrix(d),
                  axes = TRUE)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig4-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

The first example data frame is a subset of tissue gene expression data
from \citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. This data set is included in the
\CRANpkg{pald} package in an object called \texttt{tissue}.

The \texttt{tissue} object is a matrix with 189 rows, each with a
corresponding tissue, such as \texttt{colon}, \texttt{kidney} or
\texttt{cerebellum}. There are 22,215 columns corresponding to gene
expression data from each of these rows.

We will first calculate the contribution matrix for this data frame.

\begin{Schunk}
\begin{Sinput}
tissue_contribution_matrix <- tissue %>%
  get_distance_matrix(scale = FALSE) %>%
  get_contribution_matrix()
\end{Sinput}
\end{Schunk}

We can then use this contribution matrix to display the relationship
between tissue samples using the \texttt{plot\_pald\_graph()} function
(Figure \ref{fig:fig5}).

\begin{Schunk}
\begin{Sinput}
plot_pald_graph(tissue_contribution_matrix)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of tissue data]{PaLD clustering of tissue data}\label{fig:fig5}
\end{figure}
\end{Schunk}

The \texttt{get\_clusters()} function can be used to identify the
clusters of each tissue sample. Since the output is a data frame, we can
summarize the clusters using commonly used data analysis techniques. For
demonstration purposes, we will use the \CRANpkg{dplyr} package to
summarize the contribution of clusters.

\begin{Schunk}
\begin{Sinput}
library(dplyr)
get_clusters(tissue_contribution_matrix) %>%
  group_by(cluster, point) %>%
  count()
\end{Sinput}
\begin{Soutput}
#> # A tibble: 19 x 3
#> # Groups:   cluster, point [19]
#>    cluster point           n
#>      <dbl> <chr>       <int>
#>  1       1 endometrium    15
#>  2       1 kidney         39
#>  3       2 hippocampus    31
#>  4       3 cerebellum     26
#>  5       4 cerebellum      1
#>  6       5 colon          33
#>  7       6 colon           1
#>  8       7 liver           7
#>  9       8 cerebellum      1
#> 10       9 liver          17
#> 11      10 cerebellum      2
#> 12      11 liver           2
#> 13      12 cerebellum      1
#> 14      13 cerebellum      4
#> 15      14 cerebellum      2
#> 16      15 cerebellum      1
#> 17      16 placenta        2
#> 18      17 placenta        1
#> 19      18 placenta        3
\end{Soutput}
\end{Schunk}

@@

The \CRANpkg{pald} includes two data frames generated from the
scikit-learn Python package \citep{pedregosa2011scikit},
\texttt{noisy\_moons} and \texttt{noisy\_circles}.

The \texttt{noisy\_moons} data frame consists of 500 rows and two
columns.

\begin{Schunk}
\begin{Sinput}
moons_contribution_matrix <- noisy_moons %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
\end{Sinput}
\end{Schunk}

When plotting the \texttt{noisy\_moons} PaLD graph, we want the layout
to match the layout of the original data, so we will pass
\texttt{as.matrix(noisy\_moons)} to the \texttt{layout} parameter in the
\texttt{plot\_pald\_graph()} function. Additionally, here the row names
are meaningless, they just correspond to the location of the generated
data, so we can remove the labels on the plot by passing
\texttt{vertex.label\ =\ NA} to the \texttt{...} of the
\texttt{plot\_pald\_graph()} function.

\begin{Schunk}
\begin{Sinput}
plot_pald_graph(moons_contribution_matrix,
                layout = as.matrix(noisy_moons),
                vertex.label = NA)
\end{Sinput}
\end{Schunk}

The \texttt{noisy\_circles} data frame consists of 500 rows and two
columns. We can create the contribution matrix for this data frame using
the same methods as used for the \texttt{noisy\_moons} data.

\begin{Schunk}
\begin{Sinput}
circles_contribution_matrix <- noisy_circles %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
\end{Sinput}
\end{Schunk}

Similarly, we will pass the layout and remove vertex labels for the
noisy circles PaLD plot.

\begin{Schunk}
\begin{Sinput}
plot_pald_graph(circles_contribution_matrix,
                layout = as.matrix(noisy_circles),
                vertex.label = NA)
\end{Sinput}
\end{Schunk}

Because these are \CRANpkg{igraph} objects, they can be combined as you
would normally combine a plot. For example, we could add the line
\texttt{par(mfrow\ =\ c(1,\ 2))} above the plot calls to create an
output where both plots are displayed side by side (Figure
\ref{fig:fig6}).

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 2))
plot_pald_graph(moons_contribution_matrix,
                layout = as.matrix(noisy_moons),
                vertex.label = NA)
plot_pald_graph(circles_contribution_matrix,
                layout = as.matrix(noisy_circles),
                vertex.label = NA)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig6-1} \caption[PaLD clustering of noisy moons (left) and noisy circles (right) data]{PaLD clustering of noisy moons (left) and noisy circles (right) data}\label{fig:fig6}
\end{figure}
\end{Schunk}

@@

\address{%
Lucy D'Agostino McGowan\\
Wake Forest University\\
Winston-Salem, NC\\ 27106\\
}
\href{mailto:mcgowald@wfu.edu}{\nolinkurl{mcgowald@wfu.edu}}

\address{%
Katherine Moore\\
Wake Forest Unversity\\
Winston-Salem, NC\\ 27106\\
}
\href{mailto:mooreke@wfu.edu}{\nolinkurl{mooreke@wfu.edu}}

\address{%
Kenneth Berenhaut\\
Wake Forest University\\
Winston-Salem, NC\\ 27106\\
}
\href{mailto:berenhks@wfu.edu}{\nolinkurl{berenhks@wfu.edu}}

>>>>

% !TeX root = RJwrapper.tex
\title{Partitioned Local Depth (PaLD) Clustering Analyses in R}
\author{by Lucy D'Agostino McGowan, Katherine Moore, and Kenneth Berenhaut}

@@

@@

\begin{itemize}
\tightlist
\item
  Describe PaLD (Ken \& Kate?) \citep{berenhaut2022social}
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Helper functions to organize data into the correct format, into
  distance matrices and then cohesion matrices
\item
  Functions that convert a cohesion matrix into a variety of useful
  formats, including partitioned local depths, clusters, and graphs
\item
  Plotting functions
\end{enumerate}

@@

While it is not a necessity, the \CRANpkg{pald} package is designed to
function well with the pipe operator, \texttt{\textbar{}\textgreater{}}.
This functionality will be demonstrated below.

@@

For demonstration purposes, below is a sample data frame with two
variables, \texttt{x1} and \texttt{x2}. The methods put forth here work
on data frames with higher dimensions, as described in the
\textbf{Examples} section; we are simply choosing a small data frame
here for demonstration purposes.

\begin{Schunk}
\begin{Sinput}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4),
  x2 = c(5, 4, 13, 7, 18)
)
rownames(df) <- c("A", "B", "C", "D", "E")
\end{Sinput}
\end{Schunk}

The first step needed to calculate the partitioned local depths is to
construct a \emph{distance matrix}. If the data are already in this
form, the user can skip to the next step. The \texttt{dist()} function
converts an input data frame into a distance matrix, as demonstrated
below.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
\end{Sinput}
\end{Schunk}

This will create an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 5\). This distance matrix can then be passed to the
\texttt{cohesion\_matrix()} function in order to calculate the pairwise
cohesion values. Cohesion is an interpretable probability that reflects
the strength of alignment of two points. Again, if the user begins with
a distance matrix, they can skip the first step and simply input the
distance matrix into this function.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
cohesion_matrix(d)
\end{Sinput}
\begin{Soutput}
#>        A      B         C         D         E
#> A 0.2875 0.1625 0.0000000 0.0500000 0.0000000
#> B 0.1625 0.2875 0.0000000 0.0500000 0.0000000
#> C 0.0000 0.0000 0.3083333 0.1833333 0.1000000
#> D 0.0500 0.0500 0.1750000 0.3000000 0.0000000
#> E 0.0000 0.0000 0.1000000 0.0000000 0.2333333
\end{Soutput}
\end{Schunk}

Equivalently, the user can use the native pipe
\texttt{\textbar{}\textgreater{}} as follows.

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\begin{Soutput}
#>        A      B         C         D         E
#> A 0.2875 0.1625 0.0000000 0.0500000 0.0000000
#> B 0.1625 0.2875 0.0000000 0.0500000 0.0000000
#> C 0.0000 0.0000 0.3083333 0.1833333 0.1000000
#> D 0.0500 0.0500 0.1750000 0.3000000 0.0000000
#> E 0.0000 0.0000 0.1000000 0.0000000 0.2333333
\end{Soutput}
\end{Schunk}

The \emph{cohesion matrix} output by the \texttt{cohesion\_matrix()} is
the main input for the majority of the remaining functions.

\hypertarget{functions-that-convert-a-cohesion-matrix-into-useful-formats}{%
\subsubsection{Functions that convert a cohesion matrix into useful
formats}\label{functions-that-convert-a-cohesion-matrix-into-useful-formats}}

From the \emph{cohesion matrix}, a variety of useful quantities can be
calculated. Below, we create a cohesion matrix using the functions
described in the previous section.

\begin{Schunk}
\begin{Sinput}
d |>
  dist() |>
  cohesion_matrix() -> cohesion
\end{Sinput}
\end{Schunk}

To calculate the \emph{clusters} that each point will fall into, we can
use the \texttt{community\_clusters()} function. This will output a data
frame with two columns, the first will correspond to the \texttt{point},
as identified by the row name of the original input data frame,
\texttt{df}, the second will identify the \texttt{cluster} that each
point belongs to.

\begin{Schunk}
\begin{Sinput}
community_clusters(cohesion)
\end{Sinput}
\begin{Soutput}
#>   point cluster
#> A     A       1
#> B     B       1
#> C     C       2
#> D     D       2
#> E     E       3
\end{Soutput}
\end{Schunk}

@@

The \texttt{local\_depths()} function calculates the \emph{depths} of
each point, outputting a vector of local depths. Local depth is an
interpretable probability which reflects aspects of relative position
and centrality via distance comparisons (i.e., \(d(z, x) < d(z, y)\)).

\begin{Schunk}
\begin{Sinput}
local_depths(cohesion)
\end{Sinput}
\begin{Soutput}
#>         A         B         C         D         E 
#> 0.4062500 0.5416667 0.3833333 0.4687500 0.7000000
\end{Soutput}
\end{Schunk}

@@

The \texttt{strong\_threshold()} function will calculate the cohesion
threshold for strong ties. This is equal to half the average of the
diagonal cohesion matrix. This is a threshold that may be used to
distinguish between strong and weak ties.

\begin{Schunk}
\begin{Sinput}
strong_threshold(cohesion)
\end{Sinput}
\begin{Soutput}
#> [1] 0.14
\end{Soutput}
\end{Schunk}

In this case, the threshold is \texttt{0.14}.

@@

\begin{Schunk}
\begin{Sinput}
any_isolated(cohesion)
\end{Sinput}
\end{Schunk}

Here, there are no isolated points.

The function \texttt{cohesion\_strong()} will update the cohesion matrix
to set all weak ties to zero (via the \texttt{strong\_threshold()}
function). Optionally, the matrix will also be symmetrized, with the
default parameter \texttt{symmetric\ =\ TRUE}.

\begin{Schunk}
\begin{Sinput}
cohesion_strong(cohesion)
\end{Sinput}
\begin{Soutput}
#>        A      B     C     D      E
#> A 0.3000 0.1625 0.000 0.000 0.0000
#> B 0.1625 0.2875 0.000 0.000 0.0000
#> C 0.0000 0.0000 0.300 0.175 0.0000
#> D 0.0000 0.0000 0.175 0.300 0.0000
#> E 0.0000 0.0000 0.000 0.000 0.2125
\end{Soutput}
\end{Schunk}

Finally, the \texttt{community\_graphs()} function takes the cohesion
matrix and creates \CRANpkg{igraph} objects, graphs that describes the
relationship between the points. This function will output a list of
three objects:

\begin{itemize}
\tightlist
\item
  \texttt{G}: the weighted (community) graph whose edge weights are
  mutual cohesion
\item
  \texttt{G\_strong}: the weighted (community) graph consisting of edges
  for which mutual cohesion is greater than the threshold for strong
  ties
\item
  \texttt{layout}: the graph layout, using the Fruchterman Reingold (FR)
  force-directed graph drawing for the graph \texttt{G}
\end{itemize}

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 261f39e UNW- 5 2 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 261f39e (vertex names):
#> [1] A--B C--D
\end{Soutput}
\end{Schunk}

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{df} (Figure
\ref{fig:fig1}).

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `df`]{Visualize the points from data frame `df`}\label{fig:fig1}
\end{figure}
\end{Schunk}

We can then pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}).

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs()
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`]{PaLD graph displaying the relationship between the points in data frame `df`}\label{fig:fig2}
\end{figure}
\end{Schunk}

The \texttt{layout} argument allows the user to pass a matrix to dictate
the layout of the graph. For example, if we wanted the graph to match
the visualization displayed in Figure \ref{fig:fig1}, we can pass
\texttt{as.matrix(df)}, or a matrix of the data frame \texttt{df} to the
\texttt{layout} argument (Figure \ref{fig:fig3}.

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

This \texttt{plot\_community\_graphs()} function will also permit
parameters that can be passed to the \texttt{plot.igraph()} function.
For example, to add axes to the graph, the user can pass the
\texttt{axes\ =\ TRUE} argument to the \texttt{...} in the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}).

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df),
                        axes = TRUE)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig4-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

The first example if from a subset of tissue gene expression data from
\citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. A cohesion matrix was created using
this data set and is included the \CRANpkg{pald} package in an object
called \texttt{tissue\_cohesion\_matrix}.

The \texttt{tissue\_cohesion\_matrix} object is a cohesion matrix with
189 rows and 189 columns.

We can use this contribution matrix to display the relationship between
tissue samples using the \texttt{plot\_community\_graphs()} function
(Figure \ref{fig:fig5}). For clarity of the display, we show how to only
include a random set of the vertex labels. We can pass this random set
through the \texttt{...} to the \texttt{plot.igraph}
\texttt{vertex.label} parameter.

\begin{Schunk}
\begin{Sinput}
set.seed(1)

labels <- rownames(tissue_cohesion_matrix)
labels[sample(1:189, 125)] <- ""
plot_community_graphs(tissue_cohesion_matrix,
                      vertex.label = labels,
                      vertex.size = 4)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of tissue data]{PaLD clustering of tissue data}\label{fig:fig5}
\end{figure}
\end{Schunk}

The \texttt{community\_clusters()} function can be used to identify the
clusters of each tissue sample. Since the output is a data frame, we can
summarize the clusters using commonly used data analysis techniques. For
demonstration purposes, we will use the \CRANpkg{dplyr} package to
summarize the contribution of clusters.

\begin{Schunk}
\begin{Sinput}
library(dplyr)
community_clusters(tissue_cohesion_matrix) %>%
  group_by(cluster, point) %>%
  count()
\end{Sinput}
\begin{Soutput}
#> # A tibble: 19 x 3
#> # Groups:   cluster, point [19]
#>    cluster point           n
#>      <dbl> <chr>       <int>
#>  1       1 endometrium    15
#>  2       1 kidney         39
#>  3       2 hippocampus    31
#>  4       3 cerebellum     26
#>  5       4 cerebellum      1
#>  6       5 colon          33
#>  7       6 colon           1
#>  8       7 liver           7
#>  9       8 cerebellum      1
#> 10       9 liver          17
#> 11      10 cerebellum      2
#> 12      11 liver           2
#> 13      12 cerebellum      1
#> 14      13 cerebellum      4
#> 15      14 cerebellum      2
#> 16      15 cerebellum      1
#> 17      16 placenta        2
#> 18      17 placenta        1
#> 19      18 placenta        3
\end{Soutput}
\end{Schunk}

@@

The \CRANpkg{pald} includes three randomly generated data frames
corresponding to plots from \citet{berenhaut2022social}:

\begin{itemize}
\tightlist
\item
  \texttt{exdata1} is a data set consisting of 8 points to recreate
  Figure 1 in \citet{berenhaut2022social}
\item
  \texttt{exdata2} is a data set consisting of 16 points to recreate
  Figure 2 in \citet{berenhaut2022social}
\item
  \texttt{exdata3} is a data set consisting of 240 points to recreate
  Figure 4D in \citet{berenhaut2022social}
\end{itemize}

Here, we will demonstrate how to use \texttt{exdata3}. These points were
generated from bivariate normal distributions with varying means and
variances.

\begin{Schunk}
\begin{Sinput}
exdata_cohesion_matrix <- exdata3 |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\end{Schunk}

When plotting the \texttt{exdata3} PaLD graph, we want the layout to
match the layout of the original data, so we will pass
\texttt{as.matrix(exdata3)} to the \texttt{layout} parameter in the
\texttt{plot\_community\_graphs()} function. Additionally, here the row
names are meaningless, they just correspond to the location of the
generated data, so we can remove the labels on the plot by setting
\texttt{show\_labels\ =\ FALSE}. The \texttt{only\_strong} parameter
will only display the strongly connected edges. Additionally, we can
optionally pass parameters through the \texttt{...} to
\texttt{plot::igraph} function; for example, to increase the vertex size
we can set \texttt{vertex.size\ =\ 5}. Figure \ref{fig:fig6} shows this.

\begin{Schunk}
\begin{Sinput}
plot_community_graphs(exdata_cohesion_matrix,
                layout = as.matrix(exdata3),
                show_labels = FALSE, 
                only_strong = TRUE,
                vertex.size = 5)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig6-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022))}\label{fig:fig6}
\end{figure}
\end{Schunk}

@@

\address{%
Lucy D'Agostino McGowan\\
Wake Forest University\\%
Winston-Salem, NC\\ 27106\\
%
%
%
\href{mailto:mcgowald@wfu.edu}{\nolinkurl{mcgowald@wfu.edu}}%
}

\address{%
Katherine Moore\\
Wake Forest Unversity\\%
Winston-Salem, NC\\ 27106\\
%
%
%
\href{mailto:mooreke@wfu.edu}{\nolinkurl{mooreke@wfu.edu}}%
}

\address{%
Kenneth Berenhaut\\
Wake Forest University\\%
Winston-Salem, NC\\ 27106\\
%
%
%
\href{mailto:berenhks@wfu.edu}{\nolinkurl{berenhks@wfu.edu}}%
}
